---
title: "Sample Size Calculation"
subtitle: "Group 16"
author: "Valeriia Kantsypa, Bogdan-Gabriel Șlencu, Mikita Bisliuk, Silas Ooko"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
required_pkgs <- c("dplyr", "tidyr", "ggplot2", "lmerTest", "knitr", "scales", "here")
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}
invisible(lapply(required_pkgs, library, character.only = TRUE))

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 4.6,
  fig.align = "center",
  out.width = "92%",
  dpi = 150
)

options(dplyr.summarise.inform = FALSE, scipen = 999)
set.seed(20260407)

find_input_file <- function(filename) {
  candidates <- unique(c(
    filename,
    file.path(getwd(), filename),
    if (!is.null(knitr::current_input())) file.path(dirname(normalizePath(knitr::current_input())), filename),
    file.path("/mnt/data", filename)
  ))
  hit <- candidates[file.exists(candidates)][1]
  if (is.na(hit) || is.null(hit)) {
    stop(
      paste0(
        "Could not find '", filename, "'. Checked: ",
        paste(candidates, collapse = ", ")
      ),
      call. = FALSE
    )
  }
  hit
}
```

```{r calculations, include=FALSE, cache=TRUE}
pilot <- read.csv(here("data/G16.pilot.data.csv")) %>% 
  mutate(across(-quality_index, factor))

m1 <- lmer(
  quality_index ~ species + (1 | tree_id),
  data = pilot %>% filter(location == "BE")
)

pear_counts <- count(pilot, tree_id)$n
mean_pears <- mean(pear_counts)
sd_pears <- sd(pear_counts)

intercept   <- fixef(m1)["(Intercept)"]
species_eff <- fixef(m1)["speciesDoyenne"]
sd_tree     <- as.data.frame(VarCorr(m1))$sdcor[1]
sd_residual <- as.data.frame(VarCorr(m1))$sdcor[2]

climate_eff   <- 10
alpha         <- 0.05
target_power  <- 0.80
n_simulations <- 1000
ssizes        <- seq(8, 72, 8)
dropout_rate  <- 0.10

run_sim <- function(n_trees) {
  n_pears_per_tree <- pmax(1, round(rnorm(n_trees, mean = mean_pears, sd = sd_pears)))

  tree_data <- data.frame(
    tree_id = seq_len(n_trees),
    species = c(rep("Conference", n_trees / 2), rep("Doyenne", n_trees / 2)),
    climate = factor(rep(seq(4), n_trees / 4))
  ) %>%
    mutate(tree_intercept = rnorm(n(), mean = 0, sd = sd_tree))

  pear_data <- tree_data %>%
    tidyr::uncount(weights = n_pears_per_tree) %>%
    mutate(
      epsilon = rnorm(n(), mean = 0, sd = sd_residual),
      quality_index = intercept +
        (species == "Doyenne") * species_eff +
        (climate == 2) * climate_eff +
        tree_intercept +
        epsilon
    )

  m <- lmer(
    quality_index ~ species + climate + (1 | tree_id),
    data = pear_data
  )

  coef_table <- summary(m)$coefficients
  p_vals <- coef_table[grep("climate", rownames(coef_table)), "Pr(>|t|)"]
  p_vals_adj <- p.adjust(p_vals, method = "BH")

  any(p_vals_adj < alpha)
}

power <- sapply(ssizes, function(n) mean(replicate(n_simulations, run_sim(n))))

power_tbl <- data.frame(
  total_trees = ssizes,
  power = power
) %>%
  mutate(
    meets_target = power >= target_power,
    power_label = scales::percent(power, accuracy = 0.1)
  )

recommended_trees <- power_tbl %>%
  filter(meets_target) %>%
  slice(1) %>%
  pull(total_trees)

if (length(recommended_trees) == 0) {
  recommended_trees <- NA_integer_
}

recommended_with_dropout <- if (is.na(recommended_trees)) NA_integer_ else ceiling(recommended_trees / (1 - dropout_rate))

plot_obj <- ggplot(power_tbl, aes(total_trees, power)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_hline(yintercept = target_power, linetype = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_x_continuous(breaks = ssizes) +
  labs(
    x = "Total number of trees",
    y = "Estimated power"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    panel.grid.minor = element_blank()
  )

pilot_be <- pilot %>% filter(location == "BE")

param_tbl <- data.frame(
  Quantity = c(
    "Intercept",
    "Species effect: Doyenne vs Conference",
    "Tree-level SD",
    "Residual SD",
    "Mean pears per tree",
    "SD pears per tree"
  ),
  Estimate = round(c(intercept, species_eff, sd_tree, sd_residual, mean_pears, sd_pears), 2)
)

show_tbl <- power_tbl %>%
  transmute(
    `Total trees` = total_trees,
    `Estimated power` = power_label,
    `Meets 80% target` = ifelse(meets_target, "Yes", "No")
  )
```

## Objective

The **goal:** is to choose a sample size large enough to detect a **10-point quality score effect** in at least one climatic condition

**Study constraints from the analysis file**

- 2 pear species
- 4 climate settings
- 2 possible locations
- 6 Maximum number of trees per ecotron
- Preferably the study is carried out solely in Belgium
- Maximum of 12 ecotrons available
- Therefore, the cost-effective design is limited to **72 trees**
- Number of pears per tree is variable and cannot be controlled
- Expected dropout rate: **10%**

## Why simulation was used

The chosen approach was **simulation-based power analysis**, informed by the pilot data.

A simple closed-form sample size formula was not appropriate because:

- pears are nested within trees,
- trees differ from each other,
- the number of pears per tree varies,
- climate is represented by several model coefficients.

The analysis therefore keeps the planned data-generating structure and estimates power empirically.

## Pilot model used for parameter extraction

The pilot model fitted in the calculation file was:

$$
quality\_index_{ij} = \beta_0 + \beta_1 species_i + b_i + \varepsilon_{ij}
$$

with

$$
b_i \sim N(0, \sigma^2_{tree}), \qquad \varepsilon_{ij} \sim N(0, \sigma^2_{residual})
$$

This model was fitted on the **Belgium subset** of the pilot data.

```{r pilot_model_text, results='asis'}
cat(
  paste0(
    "<p><strong>Belgium pilot observations:</strong> ", nrow(pilot_be), "</p>",
    "<p><strong>Belgium pilot trees:</strong> ", dplyr::n_distinct(pilot_be$tree_id), "</p>"
  )
)
```

## Quantities taken from the pilot model

```{r parameter_table}
knitr::kable(param_tbl, align = c("l", "r"))
```

We take these values from the model calculation workflow then used in the simulation.

## How each simulated dataset was generated

For a chosen total number of trees, the simulation followed the same steps as the original calculation.

1. Number of pears per tree:

$$
N_i \sim \max\left(1, round\left(N(\mu_{pears}, \sigma^2_{pears})\right)\right)
$$

2. Random intercept for each tree:

$$
b_i \sim N(0, \sigma^2_{tree})
$$

3. Pear-level outcome:

$$
Y_{ij} = \beta_0 + I(species_i = Doyenne)\beta_{species} + I(climate_i = 2)\beta_{climate} + b_i + \varepsilon_{ij}
$$

with

$$
\varepsilon_{ij} \sim N(0, \sigma^2_{residual})
$$

and the target effect fixed at

$$
\beta_{climate} = 10
$$

## Model refitted in every simulation

Each simulated dataset was analysed with:

$$
quality\_index \sim species + climate + (1 \mid tree\_id)
$$

Then:

- p-values for climate terms were extracted,
- Benjamini-Hochberg adjustment was applied,
- power was defined as the probability that **at least one adjusted climate p-value** was below $\alpha = 0.05$.

The empirical power estimate was:

$$
\widehat{Power} = \frac{1}{R} \sum_{r=1}^{R} I(\text{at least one adjusted climate p-value is significant})
$$

with $R = 1000$ simulations for each tested sample size.

## Sample sizes evaluated

The existing calculation tested the following total numbers of trees:

$$
8,\ 16,\ 24,\ 32,\ 40,\ 48,\ 56,\ 64,\ 72
$$

These numbers show total tree count.

```{r tested_sizes}
knitr::kable(data.frame(`Total trees tested` = ssizes))
```

## Results: estimated power by total number of trees

```{r power_table_slide}
knitr::kable(show_tbl, align = c("r", "r", "c"))
```

## Results: power curve

**Power to detect a 10-point climate effect using the existing simulation approach**

```{r power_plot, fig.height=4.0, out.width='88%'}
print(plot_obj)
```

## Results: recommended sample size

```{r recommendation_text, results='asis'}
if (is.na(recommended_trees)) {
  cat("<p><strong>No sample size up to 72 trees reached 80% estimated power.</strong></p>")
} else {
  cat(
    paste0(
      "<p><strong>First sample size reaching 80% power:</strong> ",
      recommended_trees,
      " total trees.</p>"
    )
  )
}

if (is.na(recommended_with_dropout)) {
  cat("<p><strong>Dropout-adjusted target:</strong> not available because the target power was not reached in the tested range.</p>")
} else {
  cat(
    paste0(
      "<p><strong>After 10% dropout inflation:</strong> ",
      recommended_with_dropout,
      " trees.</p>"
    )
  )
}
```

## Interpretation

This slide deck does **not** change the original sample size calculation.

It presents the existing workflow in slide form:

- pilot-data model fitted on Belgium trees,
- simulation of variable pears per tree,
- mixed-model analysis with climate effects,
- Benjamini-Hochberg adjusted testing,
- empirical power across total tree counts from 8 to 72.

The recommendation in the results section is therefore directly informed by the calculation already done.
