---
title: "Valeriia-main-analysis"
output: html_document
---

```{r setup}
pacman::p_load(dplyr, tidyr, ggplot2, lmerTest, ggridges, knitr, lattice, DHARMa,
               performance, see, corrplot, lmtest, car, AER)
data <- read.csv("../data/G16.outcome.data.csv") %>%
    mutate(quality_binary = if_else(quality_idx >= 65, 1, 0))
size <- read.csv("../data/G16.size.data.csv")
soil <- read.csv("../data/G16.soil.data.csv")
```

# Objectives

**Main**

Investigate the effect of 4 controlled climate scenarios on the quality of pears.

**Side**

(a)  Transform the Gaussian outcome ‘quality index’ into a binary outcome (i.e. good quality=1/poor quality=0), with a cut-off at 65, and analyze the effect of the climate conditions on this outcome in a separate model as well, taking the potential confounding related to your experiment into account.  

(b) Analyze the effect of the climate conditions on the ‘size’ of the pears over time, taking into account the potential confounding related to your experiment.

(c) Explore/Visualize the soil condition features with a multivariate method to gain insights into the effect of climate on the pear production. Further information about the soil features are attached in a separate document.

# Exploration

## Basic stats

```{r}
cat("Pears per climate")
table(data$climate)
cat("\nPears per tree summary\n")
summary(count(data, tree_id)$n)
```
```{r}
data %>% 
    group_by(climate) %>%
    summarize(
        mean = mean(quality_idx, na.rm = TRUE),
        median = median(quality_idx, na.rm = TRUE),
        sd = sd(quality_idx, na.rm = TRUE)
    ) %>% 
    kable(caption = "Quality Index summary statistics", digits = 2)
```

```{r}
par(mfrow = c(2, 3))

hist(count(data, tree_id)$n, main = "Histogram of Pears per tree", xlab = "")

hist(table(data$ecotr_id), main = "Histogram of Pears per ecotron", xlab = "")

hist(data$quality_idx, breaks = 20, xlab = "",
     main = "Histogram of Quality Index")

boxplot(quality_idx ~ climate, data, xlab = "", ylab = "Quality Index",
        main = "Quality Index per climate")

boxplot(quality_idx ~ species, data, xlab = "", ylab = "Quality Index",
        main = "Quality Index per species")

boxplot(quality_idx ~ ecotr_id, data, xlab = "", ylab = "Quality Index",
        main = "Quality Index per ecotron")
```

- The counts of pears per tree form a bell-shaped density, which is exactly matches our approach to modelling the random number of trees
- The number of fruits under different climates as well as the number of fruits yielded from an ecotron slightly varies >> test whether the number pears varies significantly between ecotrons (just out of curiosity, the main analysis won't be affected much regardless of the result)
- !CONCERNING! a spike in 'perfect quality pears' (see "Histogram if Quality Index")
- Slight variability in Quality Index between climates / species / ecotrons

## Is the number of pears different under different climates?

```{r}
ecotron_pears <- data %>% 
    group_by(ecotr_id) %>% 
    summarize(
        climate = first(climate),
        pears = n()
    )
kruskal.test(pears ~ climate, data = ecotron_pears)
pairwise.wilcox.test(ecotron_pears$pears, ecotron_pears$climate, p.adjust.method = "BH")
```

Kruskal-Wallis test indicated a sinificant difference in the number of pears grown under at least one climate. However, post-hoc pairwise Wilcoxon rank sum test with BH correction did not produce any significant results at $0.05$ significance level. Hence, the number of pears is comparable between climates.

# Main: Quality index

$$
\text{Quality Index}_{ijt} = \beta_0 + \sum_j\beta_j \cdot \text{Climate}_j + \beta_1 \cdot \text{Species} + b_t + \varepsilon_{ijt} \\
\\
b_t \sim N(0, \sigma_\text{tree}^2) \\
\varepsilon_{ijt} \sim N(0, \sigma_\text{resid}^2)
$$

```{r fig.height=16, fig.width=16}
m1 <- lmer(quality_idx ~ climate*species + (1 | tree_id), data, REML = FALSE)
summary(m1)
check_model(m1)
```

```{r fig.height=5, fig.width=7}
m1 <- lm(quality_idx ~ climate + species, data)
qqPlot(resid(m1), main = "Linear Model Residuals Q-Q Plot", envelope = 0.95)
```

```{r fig.width=16, fig.height=8}
sim_res <- simulateResiduals(m1)
plot(sim_res)
```

- QQ plot: Probability Integral Transform (PIT) residuals are in check, uniformly distributed
- KS: Probability Integral Transform (PIT) residuals are in check, uniformly distributed
- Dispersion test: no overdispersion detected
- Outlier test: more outliers than expected (counld be attributed to a perfect-score bump)
- Residual vs. Predicted plot + Levene Test: linearity and homoscedascity assumptions are in check


```{r fig.width=16, fig.height=8}
outliers <- which(sim_res$scaledResiduals < 0.025 | sim_res$scaledResiduals > 0.975)
data[outliers, ]
```

## Beta reg

```{r}
data <- data %>% mutate(quality_idx_perc = quality_idx / 100)
m1_beta <- betareg(quality_idx_perc ~ climate*species,
                   dist = "xbetax", data)
summary(m1_beta)
```

```{r}
par(mfrow = c(1, 2), oma = c(0, 0, 3, 0))
res_pearson <- residuals(m1_beta, type = "pearson")
fitted_vals <- fitted(m1_beta)
plot(fitted_vals, res_pearson,
     xlab = "Fitted values",
     ylab = "Pearson residuals")
abline(h = 0, col = "hotpink")

qqnorm(res_pearson)
qqline(res_pearson, col = "hotpink")
mtext("m1_beta diagnostics", side = 3, outer = TRUE, font = 2, cex = 1.5)

sum(residuals(m1_beta, type="pearson")^2) / df.residual(m1_beta)
```
**Interaction**
The inclusion of an interaction term lowers AIC by 3.65 points. 
```{r}
m1_beta_no_int <- update(m1_beta, . ~ climate + species)
AIC(m1_beta, m1_beta_no_int)
```

**Heteroscedasticity**
Likelihood ratio test shows no evidence that adding a precision parameter $\phi$ improves the fit. This conclusion is further supported by comparing the AIC estimators of the constant precision model and the models accounting for possible heteroscedasticity.

```{r}
m_const <- m1_beta
m_clim  <- update(m1_beta, . ~ . | climate)
m_spec  <- update(m1_beta, . ~ . | species)
m_full  <- update(m1_beta, . ~ . | climate + species)

AIC(m_const, m_clim, m_spec, m_full)
lrtest(m_const, m_clim, m_spec, m_full)
```

```{r}
plot(m1_beta, which = c(1, 4))

```

**Multicollinearity**
Adjusted GVIF is in check for all the regression terms.

```{r}
vif(m1_beta)
```

```{r}
sim_res <- simulateResiduals(m1_beta)
plot(sim_res)
```

## Tobit

```{r}
m1_tobit <- tobit(quality_idx ~ climate*species, data = data,
                  left = -Inf, right = 100)
summary(m1_tobit)
```

```{r}
sim_res <- simulateResiduals(m1_tobit)
plot(sim_res)
```


# (a) Binary quality outcome

```{r}
m2 <- glmer(quality_binary ~ climate * species + (1 | tree_id),
           data = data,
           family = binomial)
summary(m2)
```

```{r}
exp(fixef(m2))
```

```{r}
sim_res <- simulateResiduals(m2)
plot(sim_res)
```

# (b) Pear size

```{r}
size_long <- size %>%
  pivot_longer(
    cols = starts_with("week_"), 
    names_to = "week", 
    names_prefix = "week_",
    values_to = "size_val"
  ) %>%
  mutate(week = as.numeric(week))

# 2. Plot growth curves
ggplot(size_long, aes(x = week, y = size_val, color = climate, group = pear_id)) +
  geom_line(alpha = 0.8, na.rm = TRUE) + # Individual pear trajectories (semi-transparent)
  #geom_smooth(aes(group = climate), method = "loess", size = 1.2) + # Average trend per climate
  labs(
    title = "Pear Size Growth by Climate",
    x = "Week Number",
    y = "Size",
    color = "Climate Type"
  ) +
    facet_wrap(~climate) +
  theme_minimal()
```

Seem quite linear

```{r}
m3 <- lmer(size_val ~ week * climate + species + climate:species+ (1 | tree_id / pear_id), 
           data = size_long)
summary(m3)
```
```{r fig.height=16}
check_model(m3)
```

```{r fig.width=12}
plot(simulateResiduals(m3))
```

# (c) Soil features

```{r fig.height=5, fig.width=7, warning=FALSE}
library(FactoMineR)
library(factoextra)

soil_numeric <- soil[, 7:23]
res.pca <- PCA(soil_numeric, graph = FALSE, scale.unit = TRUE)
```

```{r fig.height=5, fig.width=7, warning=FALSE}
fviz_eig(res.pca, addlabels = TRUE)
```

```{r}
fviz_cos2(res.pca, choice = "var", axes = 1:2)
```

```{r fig.width=7, fig.height=5}
var_pc_corr <- get_pca_var(res.pca)$cor
var_pc_corr <- var_pc_corr[order(var_pc_corr[,1], decreasing = TRUE), ]

corrplot(var_pc_corr)
mtext("Cos2 of variables to Dims", at=-3.5, line=2.5)
```

```{r}
fviz_contrib(res.pca, choice = "var", axes = 1:2)
```

```{r fig.width=7, fig.height=5}
var_pc_contrib <- get_pca_var(res.pca)$contrib
var_pc_contrib <- var_pc_contrib[order(var_pc_contrib[,1], decreasing = TRUE), ]
corrplot(var_pc_contrib, is.corr=FALSE)
mtext("Contributions of variables to PCs", at=-4.5, line=2.5)
```


```{r fig.width=9, fig.height=9}
# 3. Visualize the 'Individual' plot colored by Climate
fviz_pca_ind(res.pca,
             geom.ind = "point", 
             col.ind = soil$climate,
             palette = "jco",
             addEllipses = TRUE,
             #ellipse.type = "confidence",
             ellipse.type = "convex",
             legend.title = "Climate")
```


```{r fig.width=9, fig.height=9}
# 4. Visualize the 'Variables' plot to see which soil features drive the differences
fviz_pca_var(res.pca, col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

```{r}
fviz_pca_biplot(res.pca, repel = TRUE, col.ind = soil$climate, palette = "jco")
```

