---
title: "sample-size-calculation"
output: html_document
---

```{r}
pacman::p_load(dplyr, faux, lmerTest)
```

# Objective

Set the sample size sufficiently big to detect a 10 point quality score effect in at least one climate setting.

**Conditions**:
- There are 2 pear species, 4 climate settings and 2 possible locations
- Each ecotron can accomodate at most 6 trees
- Preferably the study is carried out solely in Belgium, utilizing at most 12 ecotrons. Hence, the cost-effective study is limited to 72 trees
- The number of pears per tree is variable and cannot be controlled
- Expect 10% dropout rate

# Pilot data, BE model

```{r}
pilot <- read.csv("G16.pilot.data.csv") %>% 
  mutate(across(-quality_index, factor))

# Model trees grown Belgium
m1 <- lmer(
  quality_index ~ species + (1 | tree_id),
  data = pilot %>% filter(location == "BE")
)
summary(m1)
```
```{r}
# Extract parameters to model random number of pears per tree
pear_counts <- count(pilot, tree_id)$n
mean_pears <- mean(pear_counts)
sd_pears <- sd(pear_counts)
```

```{r}
# Extract effects from the pilot data model
intercept     <- fixef(m1)["(Intercept)"]
species_eff   <- fixef(m1)["speciesDoyenne"]
sd_tree       <- as.data.frame(VarCorr(m1))$sdcor[1]
sd_residual   <- as.data.frame(VarCorr(m1))$sdcor[2]

```

# Define simulation function

```{r}
run_sim <- function(n_trees){
  
  # 1. Create a vector of pear counts per tree
  n_pears_per_tree <- pmax(1, round(rnorm(n_trees, mean = mean_pears, sd = sd_pears)))
  
  # 2. Create a dataframe with a row per tree, define:
  # tree_id, species, climate, random intercept for every tree
  tree_data <- data.frame(
    tree_id = seq_len(n_trees),
    species = c(rep("Conference", n_trees / 2), rep("Doyenne", n_trees / 2)),
    climate = factor(rep(seq(4), n_trees / 4))
  ) %>% 
    mutate(
      # Add random tree intercept
      tree_intercept = rnorm(n(), mean = 0, sd = sd_tree)
    )
  
  # 3. Model data for every pear
  pear_data <- tree_data %>% 
    tidyr::uncount(weights = n_pears_per_tree) %>% 
    mutate(
      epsilon = rnorm(n(), mean = 0, sd = sd_residual),
      quality_index = intercept +
        (species == "Doyenne") * species_eff +
        (climate == 2) * climate_eff +
        tree_intercept +
        epsilon
    )
  
  # 4. Fit the model
  m <- lmer(
    quality_index ~ species + climate + (1 | tree_id),
    data = pear_data
  )
  
  # 5. Extract p values for climate terms, adjust with BH procedure
  coef_table <- summary(m)$coefficients
  p_vals <- coef_table[grep("climate", rownames(coef_table)), "Pr(>|t|)"]
  p_vals_adj <- p.adjust(p_vals, method = "BH")
  
  # 6. Return if ANY p value is significant
  any(p_vals_adj < alpha)
}
```

# Run simulations

```{r}
# Set the climate effect to be detected
climate_eff  <- 10
alpha        <- 0.05
target_power <- 0.80
n_simulations  <- 1000

ssizes <- seq(8, 72, 8)
```

```{r}
power <- sapply(ssizes, function(n) {
  mean(replicate(n_simulations, run_sim(n)))
})

# --- PLOT ---
plot(ssizes, power, type = "b", pch = 19,
     xlab = "Sample size per group",
     ylab = "Power",
     main = "Power to detect 10-point effect (BH adjusted)")
abline(h = 0.8, col = "hotpink", lwd = 2)
```


```{r}
# Set the climate effect to be detected
climate_eff   <- 10

# Extract effects from the pilot data model
intercept     <- fixef(m)["(Intercept)"]
location_eff  <- fixef(m)["locationFR"]
species_eff   <- fixef(m)["speciesDoyenne"]
sd_tree       <- as.data.frame(VarCorr(m))$sdcor[1]
sd_residual   <- as.data.frame(VarCorr(m))$sdcor[2]

# Extract parameters to model random number of pears per tree
pear_counts <- count(pilot, tree_id)$n
mean_pears <- mean(pear_counts)
sd_pears <- sd(pear_counts)

n_trees <- 12
n_pears_per_tree <- round(rnorm(n_trees, mean = mean_pears, sd = sd_pears))
tot_obs <- sum(n_pears_per_tree)

tree_data <- data.frame(
  tree_id = seq_len(n_trees),
  species = c(rep("Conference", n_trees / 2), rep("Doyenne", n_trees / 2)),
  location = c(rep("BE", 72), rep("FR", max(0, n_trees - 72)))[1:n_trees],
  climate = rep(seq(4), n_trees / 4)
) %>% 
  mutate(
    # Add random tree intercept
    tree_intercept = rnorm(n(), mean = 0, sd = sd_tree)
  )
  tidyr::uncount(weights = n_pears_per_tree) %>% 
  mutate(
    epsilon = rnorm(n(), mean = 0, sd = sd_residual),
    tree_eff = rnorm(n(), mean = 0, sd = sd_tree),
    quality_score = intercept +
      if_else(location == "FR", 1, 0) * location_eff +
      if_else(species == "Doenne", 1, 0) * species_eff +
      tree_eff +
      epsilon
  )
```


```{r}
intercept     <- 55
species_eff   <- -6.6
climate_eff   <- 10        # ← what you want to detect
sd_tree       <- 2.4
sd_residual   <- 20.2

library(lme4)

simulate_once <- function(n_trees) {
  
  n_obs_per_tree <- round(rnorm(b_trees, mean = mean_pears, sd = sd_pears))
  
  data.frame(
    location = rep(c(0, 1), ),
    
  )
 
    
  add_random(tree = n_trees) %>%
    mutate(variety = rep(c("v1", "v2"), each = 3, length = 72)) %>%
    add_random(pear = rbinom(n = 72, size = 31, prob = 0.8), .nested_in = "tree") %>% 
    mutate(qscore = rnorm(.n, mean = ))
  # --- Assign predictors at tree level ---
  climate <- sample(c(0, 1), n_trees, replace = TRUE)
  species <- sample(c(0, 1), n_trees, replace = TRUE)
  
  # --- Random tree effects ---
  tree_eff <- rnorm(n_trees, 0, sd_tree)
  
  # --- Expand to observations ---
  tree_id <- rep(1:n_trees, each = n_obs_per_tree)
  climate_obs <- rep(climate, each = n_obs_per_tree)
  species_obs <- rep(species, each = n_obs_per_tree)
  tree_eff_obs <- rep(tree_eff, each = n_obs_per_tree)
  
  # --- Generate response ---
  y <- beta_0 +
       climate_eff * climate_obs +
       species_eff * species_obs +
       tree_eff_obs +
       rnorm(n_trees * n_obs_per_tree, 0, sd_residual)
  
  dat <- data.frame(
    y = y,
    climate = factor(climate_obs),
    species = factor(species_obs),
    tree_id = factor(tree_id)
  )
  
  # --- Fit model (simpler for speed) ---
  fit <- lm(y ~ climate + species, data = dat)
  
  pval <- summary(fit)$coefficients["climate1", "Pr(>|t|)"]
  
  return(pval < 0.05)
}
```


# Old

```{r}
# If tree 
model <- lmer(
  Y ~ climate + variety + climate:variety 
    (1 | ecotron) +
    (1 | tree),  # (1 | ecotron:tree) if tree ID is unique within ecotron
  data = dat
)
```

# Simulate the data

25 = n * p
3.3 = n*p (1-p)

p = 1 - 3.3 / 25 = 0.87
n = 25 / p = 29

```{r}
add_random(ecotron = 12) %>%
  add_random(tree = 6, .nested_in = "ecotron") %>%
  mutate(variety = rep(c("v1", "v2"), each = 3, length = 72)) %>%
  add_random(pear = rbinom(n = 72, size = 31, prob = 0.8), .nested_in = "tree") %>% 
  mutate(qscore = rnorm(.n, mean = ))
```

```{r}
add_random(climate = 2) %>%
  add_random(variety = 2, .nested_in = "climate")
```
$$Y_{ijv} = \beta_0 + \beta_j\cdot \text{climate}_j + \beta_v \cdot \text{variety}_v + \beta_{jv} \cdot \text{climate}_j \cdot \text{variety}_v + \varepsilon_{ijv}$$

```{r}
lm(score ~ variety + climate + climate:variety)
```

```{r}
add_random(variety = 2) %>% 
  add_random(climate = 4) %>% 
  add_random()
```

```{r}
library(broom)

sim_power <- function(n_pears = 25, sd_pear = 5) {
  # 1. Setup Structure: 2 Ecotrons, 6 Trees each
  data <- add_random(ecotron = 2) %>%
    add_condition("ecotron", climate = c("C1", "C2")) %>%
    add_random(tree = 6, .nested_in = "ecotron") %>%
    # Split variety 3 vs 3 within each ecotron
    add_condition("tree", variety = c("V1", "V2")) %>%
    # Generate varying pear counts
    add_random(pear = rbinom(nrow(.), 31, 0.8), .nested_in = "tree")
  
  # 2. Assign Scores
  # Let's assume V1 mean = 50, V2 mean = 55 (SD = 5)
  # Climate C2 adds +10 points
  data <- data %>%
    mutate(
      base_score = ifelse(variety == "V1", 50, 55),
      clim_eff   = ifelse(climate == "C2", 10, 0),
      # Interaction: Let's assume V2 reacts 2 points better to C2
      interact   = ifelse(climate == "C2" & variety == "V2", 2, 0),
      noise      = rnorm(nrow(.), mean = 0, sd = sd_pear),
      score      = base_score + clim_eff + interact + noise
    )
  
  # 3. Run the Linear Model
  model <- lm(score ~ variety * climate, data = data)
  
  # 4. Extract P-values
  res <- tidy(model)
  # We return TRUE if the climate effect is significant (p < .05)
  return(res$p.value[res$term == "climateC2"] < 0.05)
}

# Run 1000 iterations
sim_results <- replicate(1000, sim_power())
mean(sim_results) # This is your estimated Power
```

The simulations will be performed in the following setting:
- There is 1 ecotron per climate
- Assume homogeneity in scores between climates
- Use variance and mean of scores between the two pear varieties
- Trees of both varieties must be subject to a climate, the design is balanced
- Adjust the p-values to control for FDR by applying Benjamini-Hochberg procedure



```{r}
library(tidyverse)

# --- USER INPUTS ---
sd <- 8
mu_var1 <- 70
mu_var2 <- 60
climate_eff <- 10
power <- 0.80
alpha <- 0.05
sims <- 100
ssizes <- seq(4, 50, by = 5)

power <- sapply(ssizes, function(n) {
  res <- replicate(sims, {
    
    df <- expand.grid(variety = c(1, 2), climate = seq(4)) %>%
      slice(rep(1:n(), each = n)) %>% 
      mutate(
        variety = factor(variety),
        climate = factor(climate),
        beta_j = if_else(climate == 2, climate_eff + 5, 5),  # diff = 10
        beta_v = if_else(variety == 1, mu_var1, mu_var2),
        epsilon = rnorm(nrow(.), mean = 0, sd = sd),
        score = beta_j + beta_v + epsilon
      )
    
    m <- lm(score ~ variety + climate, data = df)
    p_vals <- summary(m)$coefficients[3:5, 4]
    p_vals_adj <- p.adjust(p_vals, method = "BH")
    
    any(p_vals_adj < 0.05)
  })
  mean(res) # Proportion of successes = Power
})

# Display results
plot(ssizes, power)
abline(h = 0.8, col = "hotpink", lwd = 1.5)
```


```{r}
library(tidyverse)

# --- PARAMETERS ---
sd <- 8
mu <- 60                  # baseline score
effect_size <- 10         # effect in ONE climate
alpha <- 0.05
sims <- 100
ssizes <- seq(4, 60, by = 4)

# --- FUNCTION TO RUN ONE SIMULATION ---
run_sim <- function(n) {
  
  # balanced design
  df <- expand.grid(
    variety = c("A", "B"),
    climate = factor(1:4)
  ) %>%
    slice(rep(1:n(), each = n))
  
  # define true effects
  df <- df %>%
    mutate(
      # only climate 1 has a real effect
      true_effect = if_else(climate == "1" & variety == "A", effect_size, 0),
      score = mu + true_effect + rnorm(n(), 0, sd)
    )
  
  # fit model with interaction
  m <- lm(score ~ variety * climate, data = df)
  
  coefs <- summary(m)$coefficients
  
  # extract p-values for variety effect within each climate
  # baseline (climate 1)
  pvals <- c(
    coefs["varietyB", 4],  # baseline climate
    
    # other climates: interaction terms
    coefs["varietyB:climate2", 4],
    coefs["varietyB:climate3", 4],
    coefs["varietyB:climate4", 4]
  )
  
  # BH correction
  p_adj <- p.adjust(pvals, method = "BH")
  
  # success = detect at least one effect
  any(p_adj < alpha)
}

# --- POWER CURVE ---
power <- sapply(ssizes, function(n) {
  mean(replicate(sims, run_sim(n)))
})

# --- PLOT ---
plot(ssizes, power, type = "b", pch = 19,
     xlab = "Sample size per group",
     ylab = "Power",
     main = "Power to detect 10-point effect (BH adjusted)")
abline(h = 0.8, col = "hotpink", lwd = 2)
```

